\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}

\usepackage{amsfonts} % mathbb
\newcommand{\nn}{\mathbb{N}_n}

\usepackage{hyperref}

\begin{document}
\title{Stable roommates problem solver}
\author{Filip Bártek}
\maketitle

\section{Problem definition}

Let's have $n$ participants.
Each participant knows some of the participants,
let's call these her potential partners.
Each participant has a linear ordering of her potential partners according
to preference.

Note that a participant may or may not consider herself a potential partner,
i.e. the relation of potential partnership needn't be irreflexive.

A matching is an equivalence relation on participants that has classes of size
at most 2,
i.e. assigns each participant one or none partner.
Matching must assign a potential partner to each of the participants.

An instability in a matching is a pair of participants each of whom prefers (according to their personal preference relations) the other to their current partner.

A stable matching is a matching that doesn't admit an instability.

In stable roommates problem, given preferences of each participant,
the task is to find a stable matching.

\subsection{Perfect matching}
A perfect matching is a matching in which every participant is assigned somebody else.

Once we can solve general SRP,
we can force a perfect matching by making sure that no participant considers herself
a potential partner.

\section{Constraint model}
In this section we'll describe a model for SRP instance on $n$ participants.

We'll use the symbol $\nn$ to denote the set $\{1, \ldots, n\}$.
Each participant is uniquely identified by a number from $\nn$.

Note that the choice of variables and constraints corresponds to the capabilities
of \emph{clpfd}, a library that is used prominently in the implementation
of the solver.

\subsection{Variables}
\paragraph{Problem instance}
We'll represent an instance of SRP with $n$ participants as
a collection of preference lists $P = (P_1, \ldots, P_n)$.
$P_i$ is a preference list expressing preferences of participant $i$.
It's a $k_i$-tuple of participant identifiers (i.e. numbers from $\nn$).
All potential partners of participant $i$ are listed in $P_i$ in order of decreasing
desirability without duplicities.

\paragraph{Problem solution}
$m: \nn \rightarrow \nn$ assigns each participant her partner.

\paragraph{Auxiliary variables}
Let $s: \nn \times \nn \rightarrow \nn$ be a score function.
$s(i,j)$ represents how desirable participant $j$ is according to participant $i$.
The lower the score, the more desirable participant $j$ is.

$s$ is defined uniquely for a given instance $P$.
For every participant $i$:

\begin{itemize}
\item $j$-th potential partner is assigned score $j$
\item every participant which is not a potential partner is assigned a score $n$
\end{itemize}

As a convenience, let $\bar{s}(i)$ denote the score that participant $i$ assigns to her partner.

\subsection{Constraints}
Since the score function $s$ depends uniquely and trivially
to the problem instance $P$,
construction of $s$ from $P$ is realized
using standard Prolog code (without the use of clpfd) and as such,
I'll leave out formal definition of the corresponding constraint.

Similarly, the properties of $P$ are not enforced formally so I'll leave out their
formal definitions.
Informal description of these properties are available in the previous section.

The following constraints constrain the solution $m$ based on $P$ and $s$:

\begin{tabular}{l | l}
Constraint & Explanation \\
\hline
$\forall i. m(i) \in P_i$ & Matching satisfies potential partners \\
$\forall i,j. m(i) = j \Leftrightarrow m(j) = i$ & Matching is symmetric \\
$\forall i. \bar{s}(i) = s(i, m(i))$ & $\bar{s}$ corresponds to $s$ and $m$ \\
$\forall i,j. \bar{s}(i) \leq s(i,j) \vee \bar{s}(j) \leq s(j,i)$ & All pairs are stable
\end{tabular}

Note especially the last of these constraints
as it captures the essence of the problem, that is the requirement of stability.
An instability occurs when a pair of participants prefer each other to their partners.
The constraint ensures that in each pair of participants,
at least one of them prefers her partner (i.e.~assigns her a lower score)
to the other participant.

\section{Implementation}
I've implemented the constraint model introduced in the previous section
using SICStus Prolog and its clpfd library.
The implementation is available in the attached file \emph{srp.pl}.

\subsection{Usage}
The main entry point is the predicate srp/2.
It's typical usage is srp(Preferences, Matching),
where Preferences is a list of ordered lists of potential partners
and Matching is unbound.
If a solution of SRP instance described by Preferences exists,
it is unified with Matching.
Otherwise srp fails.

\subsection{Performance}
I've examined the performance of the solver using various problem instances
(esp.~of various sizes) and search strategies.
%TODO: Finish.

\end{document}
